Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-11-07 RMS/submission.tex
A new version using single value as computation requirement

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{P}{eriodic} task scheduling is one of the most important
topics within the field of industrial real-time systems.
%due to the large number of control systems that require cyclic activities~\cite{buttazzo2011hard}. 
A set of periodic tasks is said to be \emph{schedulable} with respect
to some scheduling algorithm if all jobs meet their
deadlines. \emph{Rate-Monotonic Scheduling} (\emph{RMS}) is a
\emph{fixed} priority scheduling algorithm for preemptive hard
real-time environments proposed by Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73}, which assigns priorities to
jobs according to the periods of the corresponding tasks: the smaller
period, the higher priority. RMS is proved to be the \emph{optimal}
fixed priority scheduling algorithm~\cite{DBLP:journals/jacm/LiuL73},
in the sense that any set of tasks, which is schedulable under
\emph{some} fixed priority scheduling algorithm, is also schedulable
with respect to RMS. It is widely used in safety-critical real-time
applications, such as vehicles and avionics, thanks to its optimality
and easiness to implement.

Liu and Layland~\cite{DBLP:journals/jacm/LiuL73} gave a sufficient
condition for the schedulability of a set of $n$ tasks scheduled by
RMS: $\displaystyle\Sigma^n_{i=1}C_i/T_i \le n(2^{1/n}-1)$, where
$C_i$ and $T_i$ are the \emph{computation (time) requirement} and the
period of task $\tau_i$, respectively. Two main directions on RMS have
been explored since then. One is to relax the assumptions on the
original RMS model, making it applicable on more systems.  For
instance,
\cite{DBLP:conf/rtss/LehoczkySS87,DBLP:journals/rts/SpruntSL89,DBLP:conf/rtss/LehoczkyR92,DBLP:journals/tc/StrosniderLS95}
allow aperiodic tasks in the scheduling,
\cite{DBLP:journals/pe/LeungW82,audsley1993deadline} generalize RMS to
be \emph{deadline-monotonic}, \cite{DBLP:journals/tc/ShaRL90} allows
resource sharing among tasks,
\cite{dhall1978real,DBLP:journals/rts/LopezGDG03,DBLP:journals/tpds/LopezDG04,DBLP:journals/tc/BaruahG03}
extend RMS on multiprocessors, and
\cite{DBLP:journals/rts/OhS94,DBLP:journals/rts/GhoshMMS98,DBLP:journals/tpds/BertossiMR99}
enhance fault-tolerance. The other direction is to generate better
schedulablity test conditions for the algorithm and its
extensions~\cite{DBLP:conf/rtss/LehoczkySD89,DBLP:conf/rtss/KuoM91,DBLP:journals/tc/BiniBB03,DBLP:journals/rts/LopezGDG03,DBLP:journals/tc/BaruahG03,gardner1999}. The
RMS algorithm is no doubt of practical importance.

@@

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
formal modeling language and analyzing tool for real-time systems, to
model a realistic implementation of RMS, which serves as a simplified
operating system within an avionic control system, and then verify
several desired properties. Based on a realistic implementation, our
model extends the standard RMS model proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation
requirements of tasks to be non-constant, and by considering overhead
and other details of the hardware platform.

@@

(A4) All overhead, such as task switching times, is ignored.

A simple example showing this algorithm is depicted in
Figure~\ref{sf:rmsalg}. However in this paper, we consider an
implementation instead of the RMS algorithm itself, thus the model
would be more complicated than this standard, ideal
setting. Assumption~(A1) will be modified because of the interrupt
mask mechanism, while (A2) and (A4) are relaxed to obtain a more
realistic analysis model.

\subsection{Real-Time Maude}
\emph{Real-Time Maude}~\cite{DBLP:journals/lisp/OlveczkyM07} is an
extension of \emph{Maude}~\cite{DBLP:journals/tcs/ClavelDELMMQ02}
which is a language and tool based on \emph{rewriting
  logic}~\cite{DBLP:journals/jlp/Meseguer12}.  It supports formal
specification and analysis of real-time systems.

@@

In this paper, we only consider Real-Time Maude's \emph{Linear
  Temporal Logic (LTL) model checker}, which analyzes whether
\emph{each} behavior satisfies a temporal logic formula. \emph{State
  propositions} are defined as terms of sort \verb|Prop|. Their
semantics is defined by conditional equations of the form $\texttt{ceq
} statePattern \texttt{ |= } prop \texttt{ = } b \texttt{ if } cond$,
with $b$ a term of sort \verb|Bool|, stating that $prop$ evaluates to
$b$ in states which are instances of $statePattern$ provided the
condition $cond$ holds. These equations together define $prop$ to hold
in all states $s$ that make $s \texttt{ |= } prop$ evaluate to
\verb|true|. A temporal logic \emph{formula} is constructed by state
propositions and temporal logic operators such as \verb|True|,
\verb|False|, \verb|~|(negation), \verb|/\|, \verb|\/|,
\verb|->|(implication), \verb|[]|(``always'') and
\verb|U|(``until''). Real-Time Maude supports both \emph{timed} and
\emph{untimed LTL model checking}. The untimed model checking command
\begin{alltt}
  (mc \(s\) |=u \(formula\) .)
\end{alltt}
checks whether the temporal logic $formula$ holds in all behaviors
starting from the initial state $s$, \emph{with no time limit}.

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic control system.
Interrupts would be triggered by, and only by, the clock every $T$,
which we call \emph{interrupt cycle}. When an interrupt request
occurs, if the system is interruptable, i.e. the interrupt mask bit
is cleared, the handler function $schedule()$ will be invoked;
otherwise $schedule()$ will be pending until the interrupt mask bit
becomes cleared.  The implementation is shown as $schedule()$ in
Figure~\ref{a:schedule}, where $taskList$ is the list of periodic
tasks to be scheduled. We assume that the list is in descending order
of priority, and both variables $taskList$ and $timer$ are global. In
this implementation, there is only one kind of interrupt, the period
$T_i$ of each task is a multiple of $T$, and the tasks are
independent, meeting assumption~(A3).

@@

In Figure~\ref{a:schedule}, the handler function $schedule()$ first
updates status of all tasks in $taskList$ via function
$updateStatus()$. This updating actually initiates tasks which should
be scheduled in the current interrupt cycle. Then $schedule()$
traverses the list to execute the ready tasks one by one, or to do a
return when encountering an interrupted\footnote{Note that the status
  \textit{INTERRUPT} indicates the task is interrupted for the moment,
  or was interrupted before but its execution is not complete yet.}
task. As for the function $updateStatus()$, it updates each task in
two steps: firstly, if the task is running, it becomes interrupted;
secondly for the task at its initiation time, if its previous job is
complete, it would be set to be ready, otherwise it misses its
deadline, producing an error. Notice that $schedule()$ is invoked only
when the interrupt request is handled, not when the interrupt is
disabled (the interrupt mask bit is set). Due to the interrupt
mask bit, its execution cannot be interrupted when it is updating
status of tasks or finding the next task to execute, however, it can
be interrupted while executing some task (Line~\ref{l:function}). This
allows the execution of $schedule()$ to be nested.

For simplicity, in the rest of this paper, we use \emph{scheduling} to
refer to the stage, from the moment when a pending interrupt request
is detected, to the moment when the first should-be-run periodic task
starts executing, i.e. Line~\ref{l:return} or~\ref{l:function} in
Figure~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when an interrupt request
is handled, (ii) the time spent by $schedule()$ finding and setting
the \emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in Figure~\ref{a:schedule}),
and (iii) the time for switching context from $schedule()$ to that
task. \emph{Switching} refers to the stage, from the moment when a
periodic task completes its execution, to the moment when the next
should-be-run periodic task starts executing. \emph{Switching time}
thus also consists of three parts: (i) the time for switching context
from the complete task back to $schedule()$, (ii) the time spent by
$schedule()$ finding and setting the \emph{next} should-be-run
periodic task, and (iii) the time for switching context from
$schedule()$ to that task.

@@

(A2') Computation requirement $C_i$ for each task $\tau_i$ is not
constant. It varies randomly between minimum $C^{min}_i$ and maximum
$C^{max}_i$, i.e. within an interval $[C^{min}_i, C^{max}_i]$.

@@

These assumptions make our model different from the standard one. For
instance, with assumption~(A1'), an interrupt request that occurs
while switching is performed will be pending, such that jobs of task
$\tau_i$ cannot initiate at $kT_i$. They should wait until switching
finishes and the interrupt mask bit is cleared, which is different
from~(A1). Note that (A3) says that jobs are ready to run at their
initiation times, however, no one can start running at exactly its
initiation time, because scheduling takes time. Under these
assumptions, an example for showing the execution of tasks scheduled
by the implementation is depicted in Figure~\ref{sf:rmsimp}. Note that
the second job of $\tau_1$ is initiated at time $5.5$ instead of $5$,
because switching is performed and the interrupt mask bit is set at
time $5$.

@@

    % time indicator
    \draw [time line] (1.5,0.7) -- (1.5,0);
    \draw (1.5,0) node [below] {$3$} -- +(0,2pt);
    \draw [time line] (2.5,0.7) -- (2.5,0);
    \draw (2.5,0) node [below] {$5$} -- +(0,2pt);
    \draw [time line] (4,0.7) -- (4,0);
    \draw (4,0) node [below] {$8$} -- +(0,2pt);
    \draw [time line] (4.5,0.3) -- (4.5,0);
    \draw (4.5,0) node [below] {$9$} -- +(0,2pt);
    \draw [time line] (5,0.7) -- (5,0);
    \draw (5,0) node [below] {$10$} -- +(0,2pt);

    % tasks
    \draw [periodic task] (0,0.7) rectangle +(1.5,0.14);
    \draw [periodic task] (1.5,0.3) rectangle +(1,0.14);
    \draw [periodic task] (2.5,0.7) rectangle +(1.5,0.14);
    \draw [periodic task] (4,0.3) rectangle +(0.5,0.14);

@@

    % completion of tasks
    \draw [complete] (1.5,0.84) -- +(0,0.01);
    \draw [complete] (4,0.84) -- +(0,0.01);
    \draw [complete] (4.5,0.44) -- +(0,0.01);
  \end{tikzpicture}
\label{sf:rmsalg}}\\
\subfloat[RMS Implementation]{
  \begin{tikzpicture}[font=\small,scale=1.25,
    % styles
    system task/.style={fill=gray!70,draw=gray},
    periodic task/.style={fill=gray!70!black,draw=gray!50!black},
    task line/.style={gray!30,very thin},
    time line/.style={gray!50,very thin,dashed},
    short line/.style={gray!50,very thin,dashed},
    init/.style={{Triangle[fill=white,scale=1.1]}-},
    complete/.style={{Triangle[scale=1.1]}-}]

@@

    % time indicator
    \draw [time line] (0.5,1.5) -- (0.5,0);
    \draw (0.5,0) node [below] {$1$} -- +(0,2pt);
    \draw [time line] (1.25,1.5) -- (1.25,0);
    \draw (1.25,0) node [below] {$2.5$} -- +(0,2pt);
    \draw [time line] (1.75,1.1) -- (1.75,0);
    \draw (1.75,0) node [below] {$3.5$} -- +(0,2pt);
    \draw [time line] (2.25,1.1) -- (2.25,0);
    \draw (2.25,0) node [below] {$4.5$} -- +(0,2pt);
    \draw [time line] (2.75,1.5) -- (2.75,0);
    \draw (2.75,0) node [below] {$5.5$} -- +(0,2pt);
    \draw [time line] (3.25,1.5) -- (3.25,0);
    \draw (3.25,0) node [below] {$6.5$} -- +(0,2pt);
    \draw [time line] (4.25,1.5) -- (4.25,0);
    \draw (4.25,0) node [below] {$8.5$} -- +(0,2pt);
    \draw [time line] (4.75,0.3) -- (4.75,0);
    \draw (4.75,0) -- +(0,2pt);
    \draw (4.65,0) node [below] {$9.5$};
    \draw [time line] (5,1.5) -- (5,0);
    \draw (5,0) node [below] {$10$} -- +(0,2pt);

    % tasks
    \draw [system task] (0,0.7) rectangle +(0.5,0.14);
    \draw [periodic task] (0.5,1.5) rectangle +(0.75,0.14);
    \draw [system task] (1.25,0.3) rectangle +(0.5,0.14);
    \draw [periodic task] (1.75,1.1) rectangle +(0.5,0.14);
    \draw [system task] (2.25,0.3) rectangle +(0.5,0.14);
    \draw [system task] (2.75,0.7) rectangle +(0.5,0.14);
    \draw [periodic task] (3.25,1.5) rectangle +(1,0.14);
    \draw [system task] (4.25,0.3) rectangle +(0.5,0.14);

@@

    % completion of tasks
    \draw [complete] (1.25,1.64) -- +(0,0.01);
    \draw [complete] (2.25,1.24) -- +(0,0.01);
    \draw [complete] (4.25,1.64) -- +(0,0.01);
  \end{tikzpicture}
\label{sf:rmsimp}}
\caption{A task set scheduled under RMS algorithm and the
  implementation, respectively. $T=T_1=5$, $T_2=10$.  In
  Figure~\ref{sf:rmsalg}, with assumption~(A2), $C_1=3$, $C_2=3$. In
  Figure~\ref{sf:rmsimp}, with assumption~(A2'), $C_1\in [1,3]$ and
  $C_2\in [1,3]$, assuming both scheduling time and switching time are
  $1$.}
\label{f:example}
\end{figure}

@@

We also need a sort \verb|Interval| to denote computation requirements
of tasks, and a sort \verb|Counter| to record execution of tasks.  We
call \emph{execution time} the time how long a task has been executing
for. A \verb|Counter| records the execution time and the computation
requirement of a task.
\begin{verbatim}
 op [_,_] : Time Time -> Interval [ctor] .  
 op [_/_] : Time Interval 
              -> Counter [ctor] .
\end{verbatim}

@@

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes:
\begin{verbatim}
 class SysTask .            
 subclass SysTask < Task .
\end{verbatim}
Different from periodic tasks, system tasks are organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.
%We abstract here from the detailed definitions thanks to the
%similarity.

@@

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Figure~\ref{a:schedule}, which is treated as an
instantaneous action in our model. It is modeled by the recursive
function as below:
\begin{verbatim}
 op updateStatus_with_ : TaskList Timer 
                           -> TaskList . 
 eq updateStatus null with TIMER = null .
 eq updateStatus (TASK :: L) with TIMER
    = (update TASK with TIMER) 
      :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and function \verb|update_with_| updating the status of individual
task (Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Figure~\ref{a:schedule}):
\begin{verbatim}
 op update_with_ : Object Timer 
                     ~> Object .
 ceq update < O : PTask | period : T, 
                          status : ST > 
       with TIMER
     = if ST == DORMANT 
       then < O : PTask | status : READY >
       else error fi
     if TIMER rem T == 0 .
 eq update < O : PTask | status : ST > 
      with TIMER
    = if ST == RUNNING 
      then < O : PTask | status 
                           : INTERRUPT >
      else < O : PTask |> fi [owise] .
\end{verbatim}
Given a task, if \verb|TIMER|($timer$) can be divided by its period,
this task should be initiated.  In the case where the task should be
initiated, it is set to be \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the \verb|mask| bit to be cleared. Once the
request is detected, it is handled in two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the $schedule()$
function. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
 crl [interrupt-handle] :
   SYSTEM 
   => ((SYSTEM).interrupt).startScheduling
   if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the \verb|ir| bit, which means the request
has been handled; (ii) pushing the current \verb|pc| value into the
stack, storing the interrupted context; (iii) assigning
\verb|scheduling| of sort \verb|Oid| to \verb|pc|, which indicates
that the system is scheduling; and (iv) setting the \verb|mask| bit,
to mask coming interrupt requests.

>>>>

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{P}{eriodic} task scheduling is one of the most important
topics within the field of industrial real-time systems.
%due to the large number of control systems that require cyclic activities~\cite{buttazzo2011hard}. 
A set of periodic tasks is said to be \emph{schedulable} with respect
to some scheduling algorithm if all jobs meet their
deadlines. \emph{Rate-Monotonic Scheduling} (\emph{RMS}) is a
\emph{fixed} priority scheduling algorithm for preemptive hard
real-time environments proposed by Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73}, which assigns priorities to
jobs according to the periods of the corresponding tasks: the smaller
period, the higher priority. RMS is proven to be the \emph{optimal}
fixed priority scheduling algorithm~\cite{DBLP:journals/jacm/LiuL73},
in the sense that any set of tasks, which is schedulable under
\emph{some} fixed priority scheduling algorithm, is also schedulable
with respect to RMS. It is widely used in safety-critical real-time
applications, such as vehicles and avionics, thanks to its optimality
and easiness to implement.

Liu and Layland~\cite{DBLP:journals/jacm/LiuL73} gave a sufficient
condition for the schedulability of a set of $n$ tasks scheduled by
RMS: $\displaystyle\Sigma^n_{i=1}C_i/T_i \le n(2^{1/n}-1)$, where
$C_i$ and $T_i$ are the \emph{computation (time) requirement} and the
period of task $\tau_i$, respectively. Two main directions on RMS have
been explored since then. One is to relax the assumptions on the
original RMS model, making it applicable on more systems.  For
instance,
\cite{DBLP:conf/rtss/LehoczkySS87,DBLP:journals/rts/SpruntSL89,DBLP:conf/rtss/LehoczkyR92,DBLP:journals/tc/StrosniderLS95}
allow aperiodic tasks in the scheduling,
\cite{DBLP:journals/pe/LeungW82,audsley1993deadline} generalize RMS to
be \emph{deadline-monotonic}, \cite{DBLP:journals/tc/ShaRL90} allows
resource sharing among tasks,
\cite{dhall1978real,DBLP:journals/rts/LopezGDG03,DBLP:journals/tpds/LopezDG04,DBLP:journals/tc/BaruahG03}
extend RMS on multiprocessors, and
\cite{DBLP:journals/rts/OhS94,DBLP:journals/rts/GhoshMMS98,DBLP:journals/tpds/BertossiMR99}
enhance fault-tolerance. The other direction is to generate better
schedulability test conditions for the algorithm and its
extensions~\cite{DBLP:conf/rtss/LehoczkySD89,DBLP:conf/rtss/KuoM91,DBLP:journals/tc/BiniBB03,DBLP:journals/rts/LopezGDG03,DBLP:journals/tc/BaruahG03,gardner1999}. The
RMS algorithm is no doubt of practical importance.

@@

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
formal modeling language and analysis tool for real-time systems, to
model a realistic implementation of RMS, which serves as a simplified
operating system within an avionic control system, and then verify
several desired properties. Based on a realistic implementation, our
model extends the standard RMS model proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by considering overhead and other
details of the hardware platform.

@@

(A4) All overhead, such as task switching time, is ignored.

A simple example showing this algorithm is depicted in
Figure~\ref{sf:rmsalg}. However in this paper, we consider an
implementation instead of the RMS algorithm itself, thus the model
would be more complicated than this standard, ideal
setting. Assumption~(A1) will be modified because of the interrupt
mask mechanism, while (A4) is relaxed to obtain a more realistic
analysis model.

\subsection{Real-Time Maude}
Real-Time Maude~\cite{DBLP:journals/lisp/OlveczkyM07} is an extension
of \emph{Maude}~\cite{DBLP:journals/tcs/ClavelDELMMQ02} which is a
language and tool based on \emph{rewriting
  logic}~\cite{DBLP:journals/jlp/Meseguer12}.  It supports formal
specification and analysis of real-time systems.

@@

In this paper, we only consider Real-Time Maude's \emph{Linear
  Temporal Logic (LTL) model checker}, which analyzes whether
\emph{each} behavior satisfies a temporal logic formula. \emph{State
  propositions} are defined as terms of sort \verb|Prop|. Their
semantics is defined by conditional equations of the form $\texttt{ceq
} statePattern \texttt{ |= } prop \texttt{ = } b \texttt{ if } cond$,
with $b$ a term of sort \verb|Bool|, stating that $prop$ evaluates to
$b$ in states which are instances of $statePattern$ provided the
condition $cond$ holds. These equations together define $prop$ to hold
in all states $s$ that make $s \texttt{ |= } prop$ evaluate to
\verb|true|. A temporal logic \emph{formula} is constructed by state
propositions and temporal logic operators such as \verb|True|,
\verb|False|, \verb|~|(negation), \verb|/\|, \verb|\/|,
\verb|->|(implication), \verb|[]|(``always'') and
\verb|U|(``until''). Real-Time Maude supports both \emph{timed} and
\emph{untimed LTL model checking}. The untimed model checking command
\begin{alltt}
  (mc \(s\) |=u \(\Phi\) .)
\end{alltt}
checks whether the temporal logic formula $\Phi$ holds in all
behaviors starting from the initial state $s$, \emph{with no time
  limit}.

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in an industrial avionic control system.
Interrupts would be triggered by, and only by, the clock every $T$,
which we call \emph{interrupt cycle}. When an interrupt request
occurs, if the system is interruptable, i.e. the interrupt mask bit is
cleared, the handler function $schedule()$ will be invoked; otherwise
$schedule()$ will be pending until the interrupt mask bit becomes
cleared.  The implementation is shown as $schedule()$ in
Figure~\ref{a:schedule}, where $taskList$ is the list of periodic
tasks to be scheduled. We assume that the list is in descending order
of priority, and both variables $taskList$ and $timer$ are global. In
this implementation, there is only one kind of interrupt, the period
$T_i$ of each task is a multiple of $T$, and the tasks are
independent, meeting assumption~(A3).

@@

In Figure~\ref{a:schedule}, the handler function $schedule()$ first
updates status of all tasks in $taskList$ via function
$updateStatus()$. This updating actually initiates tasks that should
be scheduled in the current interrupt cycle. Then $schedule()$
traverses the list to execute the ready tasks one by one, or to do a
return when encountering an interrupted\footnote{Note that the status
  \textit{INTERRUPT} indicates the task is interrupted for the moment,
  or was interrupted before but its execution is not complete yet.}
task. As for the function $updateStatus()$, it updates each task in
two steps: firstly, if the task is running, it becomes interrupted;
secondly for the task at its initiation time, if its previous job is
complete, it would be set ready, otherwise it misses its deadline,
producing an error. Notice that $schedule()$ is invoked only when the
interrupt request is handled, not when the interrupt is disabled (the
interrupt mask bit is set). Due to the interrupt mask bit, its
execution cannot be interrupted when it is updating status of tasks or
searching the next task to execute, however, it can be interrupted while
executing some task (Line~\ref{l:function}). This allows the execution
of $schedule()$ to be nested.

For simplicity, in the rest of this paper, we use \emph{scheduling} to
refer to the stage, from the moment when a pending interrupt request
is detected, to the moment when the first should-be-run periodic task
starts executing, i.e. Line~\ref{l:return} or~\ref{l:function} in
Figure~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when an interrupt request
is handled, (ii) the time spent by $schedule()$ searching and setting
the \emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in Figure~\ref{a:schedule}),
and (iii) the time for switching context from $schedule()$ to that
task. \emph{Switching} refers to the stage, from the moment when a
periodic task completes its execution, to the moment when the next
should-be-run periodic task starts executing. \emph{Switching time}
thus also consists of three parts: (i) the time for switching context
from the complete task back to $schedule()$, (ii) the time spent by
$schedule()$ searching and setting the \emph{next} should-be-run
periodic task, and (iii) the time for switching context from
$schedule()$ to that task.

@@

@@

These assumptions make our model different from the standard one. For
instance, with assumption~(A1'), an interrupt request that occurs
during switching will be pending, such that jobs of task $\tau_i$
cannot initiate at $kT_i$. They should wait until switching finishes
and the interrupt mask bit is cleared, which is different
from~(A1). Note that (A3) says that jobs are ready to run at their
initiation times, however, no one can start running at exactly its
initiation time, because scheduling takes time. Under these
assumptions, an example showing the execution of tasks scheduled by
the implementation is depicted in Figure~\ref{sf:rmsimp}. Note that
the second job instance of $\tau_1$ is initiated at time $11$ instead
of $10$, because switching is performed and the interrupt mask bit is
set at time $10$.

@@

    % time indicator
    \draw [time line] (0.75,0.7) -- (0.75,0);
    \draw (0.75,0) node [below] {$3$} -- +(0,2pt);
    \draw [time line] (1.25,0.3) -- (1.25,0);
    \draw (1.25,0) node [below] {$5$} -- +(0,2pt);
    \draw [time line] (2.5,0.7) -- (2.5,0);
    \draw (2.5,0) node [below] {$10$} -- +(0,2pt);
    \draw [time line] (3.25,0.7) -- (3.25,0);
    \draw (3.25,0) node [below] {$13$} -- +(0,2pt);

    \draw [time line] (5,0.7) -- (5,0);
    \draw (5,0) node [below] {$20$} -- +(0,2pt);

    % tasks
    \draw [periodic task] (0,0.7) rectangle +(0.75,0.14);
    \draw [periodic task] (0.75,0.3) rectangle +(0.5,0.14);
    \draw [periodic task] (2.5,0.7) rectangle +(0.75,0.14);

@@

    % completion of tasks
    \draw [complete] (0.75,0.84) -- +(0,0.01);
    \draw [complete] (3.25,0.84) -- +(0,0.01);
    \draw [complete] (1.25,0.44) -- +(0,0.01);
  \end{tikzpicture}
\label{sf:rmsalg}}\\
\subfloat[RMS Implementation]{
  \begin{tikzpicture}[font=\small,scale=1.25,
    % styles
    system task/.style={fill=gray!70,draw=gray},
    periodic task/.style={fill=gray!70!black,draw=gray!50!black},
    task line/.style={gray!30,very thin},
    time line/.style={gray!50,very thin,dashed},
    short line/.style={gray!50,very thin,dashed},
    init/.style={{Triangle[fill=white,scale=1.1]}-},
    complete/.style={{Triangle[scale=1.1]}-}]

@@

    % time indicator
    \draw [time line] (0.5,1.5) -- (0.5,0);
    \draw (0.5,0) node [below] {$2$} -- +(0,2pt);
    \draw [time line] (1.25,1.5) -- (1.25,0);
    \draw (1.25,0) node [below] {$5$} -- +(0,2pt);
    \draw [time line] (1.75,1.1) -- (1.75,0);
    \draw (1.75,0) node [below] {$7$} -- +(0,2pt);
    \draw [time line] (2.25,1.1) -- (2.25,0);
    \draw (2.25,0) node [below] {$9$} -- +(0,2pt);
    \draw [time line] (2.75,1.5) -- (2.75,0);
    \draw (2.75,0) node [below] {$11$} -- +(0,2pt);
    \draw [time line] (3.25,1.5) -- (3.25,0);
    \draw (3.25,0) node [below] {$13$} -- +(0,2pt);
    \draw [time line] (4,1.5) -- (4,0);
    \draw (4,0) node [below] {$16$} -- +(0,2pt);
    \draw [time line] (4.5,0.3) -- (4.5,0);
    \draw (4.5,0) -- +(0,2pt);
    \draw (4.5,0) node [below] {$18$};
    \draw [time line] (5,1.5) -- (5,0);
    \draw (5,0) node [below] {$20$} -- +(0,2pt);

    % tasks
    \draw [system task] (0,0.7) rectangle +(0.5,0.14);
    \draw [periodic task] (0.5,1.5) rectangle +(0.75,0.14);
    \draw [system task] (1.25,0.3) rectangle +(0.5,0.14);
    \draw [periodic task] (1.75,1.1) rectangle +(0.5,0.14);
    \draw [system task] (2.25,0.3) rectangle +(0.5,0.14);
    \draw [system task] (2.75,0.7) rectangle +(0.5,0.14);
    \draw [periodic task] (3.25,1.5) rectangle +(0.75,0.14);
    \draw [system task] (4,0.3) rectangle +(0.5,0.14);

@@

    % completion of tasks
    \draw [complete] (1.25,1.64) -- +(0,0.01);
    \draw [complete] (2.25,1.24) -- +(0,0.01);
    \draw [complete] (4,1.64) -- +(0,0.01);
  \end{tikzpicture}
\label{sf:rmsimp}}
\caption{A set of $2$ periodic tasks scheduled under RMS algorithm and
  the implementation, respectively. $T=T_1=10$, $T_2=20$, $C_1=3$,
  $C_2=2$. In Figure~\ref{sf:rmsimp}, we assume both scheduling time
  and switching time are $2$.}
\label{f:example}
\end{figure}

@@

We also need
%a sort \verb|Interval| to denote computation requirements of tasks, and 
a sort \verb|Counter| to record the execution of tasks.  We call
\emph{execution time} the time how long a task has been executing
for. A \verb|Counter| records the execution time and the computation
requirement of a task.
\begin{verbatim}
 op [_/_] : Time Time -> Counter [ctor] .
\end{verbatim}

@@

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes:
\begin{verbatim}
 class SysTask | .            
 subclass SysTask < Task .
\end{verbatim}
Different from periodic tasks, system tasks are organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.
%We abstract here from the detailed definitions thanks to the
%similarity.

@@

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Figure~\ref{a:schedule}, which is treated as an
instantaneous action in our model. It is modeled by the recursive
function as below:
\begin{verbatim}
 op updateStatus_with_ : TaskList Timer 
                           -> TaskList . 
 eq updateStatus null with TIMER = null .
 eq updateStatus (TASK :: L) with TIMER
    = (update TASK with TIMER) 
      :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and function \verb|update_with_| updating the status of individual
task (Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Figure~\ref{a:schedule}):
\begin{verbatim}
 op update_with_ : Object Timer 
                     ~> Object .
 ceq update < O : PTask | period : T, 
                          status : ST > 
       with TIMER
     = if ST == DORMANT 
       then < O : PTask | status : READY >
       else error fi
     if TIMER rem T == 0 .
 eq update < O : PTask | status : ST > 
      with TIMER
    = if ST == RUNNING 
      then < O : PTask | status 
                           : INTERRUPT >
      else < O : PTask |> fi [owise] .
\end{verbatim}
Given a task, if \verb|TIMER|($timer$) can be divided by its period,
this task should be initiated.  In the case where the task should be
initiated, it is set \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the bit \verb|mask| to be cleared. Once the
request is detected, it is handled in two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the function
$schedule()$. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
 crl [interrupt-handle] :
   SYSTEM 
   => ((SYSTEM).interrupt).startScheduling
   if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the bit \verb|ir|, which means the request
has been handled; (ii) pushing the current \verb|pc| value into the
stack, storing the interrupted context; (iii) assigning
\verb|scheduling| of sort \verb|Oid| to \verb|pc|, which indicates
that the system is scheduling; and (iv) setting the bit \verb|mask|,
to mask coming interrupt requests.
